---
title: "Reproducible Resesearch: Peer Assessment 2"
author: "Kwamena Appiah-Kubi"
date: "26 July 2015"
output: html_document
---

# Analysis of economic and population impact of storms across the United States.
Using data from the U.S. National Ocenic and Atmospheric Administration's storm databbase of major storm and weather events from 1950 to 2011, the following analysis explores which events impacts population health the most across the United States. It also investigates which events result in greatest economic consequence across th United States. For the analysis only the last decade's data will be considered (2001 - 2011)

## Data processing
```{r echo=TRUE, cache=FALSE}
# reading and reformatting the date column
storm_data <- read.csv("repdata-data-StormData.csv.bz2", stringsAsFactors = FALSE, na.strings = "")
storm_data$BGN_DATE <- as.Date(storm_data$BGN_DATE, format = "%m/%d/%Y")
storm_data$storm_yr <- format(storm_data$BGN_DATE, "%Y")

library(data.table)
storm_data <- as.data.table(storm_data)
```


## Results

```{r echo=TRUE, results='hide'}
library(ggplot2)

total_pop_impact_by_event <- storm_data[, .(total_impact = sum(FATALITIES + INJURIES)), by = c("EVTYPE")]


```
```{r echo=TRUE}
qplot(EVTYPE, data = total_pop_impact_by_event[total_impact > 100], weight = total_impact) + 
        geom_bar() + 
        ylab("Number of fatalities and injuries") +
        xlab("Weather Event type") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        ggtitle("Weather Events across U.S. with injuries & impact greather than 100 from 1995 - 2011")

```


From the plot tornada causes the most sever impact on population health, followed by exessive heat and lightning.


```{r echo=TRUE}

economic_impact_data <- storm_data[, .(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)]
economic_impact_data <- economic_impact_data[complete.cases(economic_impact_data)]
economic_impact_data[PROPDMGEXP == "K", PROPDMG := PROPDMG * 1000]
economic_impact_data[PROPDMGEXP == "M", PROPDMG := PROPDMG * 1000000]
economic_impact_data[PROPDMGEXP == "B", PROPDMG := PROPDMG * 1000000000]
economic_impact_data[CROPDMGEXP == "K", CROPDMG := CROPDMG * 1000]
economic_impact_data[CROPDMGEXP == "M", CROPDMG := CROPDMG * 1000000]
economic_impact_data[CROPDMGEXP == "B", CROPDMG := CROPDMG * 1000000000]

total_economic_impact_by_event <- economic_impact_data[, .(total_impact = sum(CROPDMG + PROPDMG)), by = c("EVTYPE")]
total_economic_impact_by_event <- total_economic_impact_by_event[total_impact > 1000000]

total_economic_impact_by_event[, total_impact := total_impact / 1000000]

qplot(EVTYPE, data = total_economic_impact_by_event, weight = total_impact) + 
        geom_bar() + 
        ylab("Total value in $1M") +
        xlab("Weather Event type") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        ggtitle("Economic impact greater than $1M of Weather Events across U.S. from 1995 - 2011")

```


From the plot, floods cause the most damaging economic impact followed by hurricanes/typhoons and tornados. 